\chapter{Conclusions}\label{ch:conclusion}

In this study, we investigated some decidability questions in the
setting of abstract interpretation, the most prevalent framework for
developing static analysis tools.  The general question concerned the
possibility of effectively computing the abstract semantics. The
question is non trivial when the abstract domain can have infinite
ascending chains and thus the fixpoint computations can diverge.  We
focused on a simple imperative language \(\imp\), offering a set of
constructs sufficient to make it Turing complete. We defined its
concrete denotational semantics in its concrete domain \(\dom\) and
small step semantics. We first showed that, as expected, all
non-trivial properties of the collecting semantics are undecidable due
to Rice's Theorem. We then built the abstraction chain in
Figure~\ref{fig:abstrchain}.
\begin{figure}
  \centering
  \usetikzlibrary{arrows.meta}
  \begin{tikzpicture}[, >=stealth]
    % Nodes
    \node[minimum size=20pt] (dom) {\(\dom\)};
    \node[minimum size=20pt] (bcnr) [below=of dom] {\(\bCnr\)};
    \node[minimum size=20pt] (ph1) [below=of bcnr] {};
    \node[minimum size=20pt] (inte) [left=of ph1] {\(\inte\)};
    \node[minimum size=20pt] (bbcnr) [right=of ph1] {\(\bbCnr{\low}{\upp}\)};
    \node[minimum size=20pt] (binte) [below=of inte] {\(\binte{\low}{\upp}\)};
    \node[minimum size=20pt] (btbcnr) [below=of bbcnr] {\(\btbCnr{\low}{\upp}\)};

    \node at (-4,-3) (decidable)   {{\color{codegreen}\(\mathsf{decidable}\)}};
    \node at (3, -2) (unknown)     {{\color{airforceblue}\(\mathsf{unknown}\)}};
    \node at (2, .5) (undecidable) {{\color{red}\(\mathsf{undecidable}\)}};

    % % divisor lines
    \draw[red, opacity=.4] (-3,1) edge[out=-40,in=180, dashed] (3,-1);
    \draw[airforceblue, opacity=.4] (-3,0.9) edge[out=-40,in=180, dashed] (3,-1.1);
    % % \draw (description) edge[out=180,in=0,->] (text);
    \draw[airforceblue, opacity=.4] (-3,-1) edge[out=-50,in=180, dashed] (3,-4.2);
    \draw[codegreen, opacity=.4] (-3,-1.1) edge[out=-50,in=180, dashed] (3,-4.3);
    
    % Arrows
    \path
    (dom.250) edge[->] node[left]  {$\abstr$}             (bcnr.110)
    (bcnr.70) edge[->] node[right] {$\concr$}             (dom.290)
    (bcnr.195)    edge[->] node[left]  {$\dabstr$}        (inte.75)
    (inte.35)    edge[->] node[right] {$\dconcr$}         (bcnr.235)
    (bcnr.305)    edge[->] node[left]  {$\sabstr[\low,\upp]$} (bbcnr.145)
    (bbcnr.115)   edge[->] node[right] {$\sconcr[\low,\upp]$} (bcnr.335)
    (inte.250)    edge[->] node[left]  {$\dabstr[\low,\upp]$} (binte.110)
    (binte.70)   edge[->] node[right] {$\dconcr[\low,\upp]$} (inte.290)
    (bbcnr.250)   edge[->] node[left]  {$\tabstr[\low,\upp]$} (btbcnr.110)
    (btbcnr.70)  edge[->] node[right] {$\tconcr[\low,\upp]$} (bbcnr.290);
    
    % % Decidable set
    % \filldraw[fill=green, fill opacity=0.1, draw=green, dotted, rounded corners] 
    % (inte.north -| inte.west) to[out=210, in=120]
    % (binte.south -| binte.west) to[out=330, in=225] 
    % (btbcnr.south -| btbcnr.east) to[out=45, in=315]
    % (btbcnr.north -| btbcnr.east) to[out=135, in=-10] 
    % (inte.north -| inte.east) to[out=170, in=10]
    % (inte.north -| inte.west);

    % % Unknown set
    % \filldraw[fill=airforceblue, fill opacity=0.1, draw=airforceblue, dotted] 
    % (bcnr.south -| bcnr.west) to[out=315, in=155]
    % (bbcnr.south -| bbcnr.west) to[out=325, in=315]
    % (bbcnr.north -| bbcnr.east) to[out=135, in=325]
    % (bcnr.north -| bcnr.east) to[out=155, in=135]
    % (bcnr.south -| bcnr.west);

  \end{tikzpicture}
  \caption{Abstractions chain we build throughout
    Chapter~\ref{ch:abstractdomains}}\label{fig:abstrchain}
\end{figure}
First we abstracted program execution using the non-relational
collecting semantics \(\bCnr\), the finest non-relational
domain. Subsequently, we defined and established the fundamental
properties of the intervals domain \(\inte\), a well-known domain in
abstract interpretation literature. We proved that in the
aforementioned abstract domains, variables are constrained to increase
up to a limit denoted as \(\bound{\com}\), which is dependent on the
program \(\com\). We therefore introduced the bounded interval domain
\(\binte{\low}{\upp}\): such domain does not contain infinite
ascending and descending chains, and therefore fixpoints are
computable in finite time. We then proved that given a program and an
initial environment \(\com\in\imp, \eta\in\inte\) and by crafting
appropriate bounds for \(\binte{\low}{\upp}\), namely
\(\low = \min(\eta) - \nlbound{\com}\) and
\(\upp = \max(\eta) + \nbound{\com}\), we not only could compute
\(\semi[\binte{\low}{\upp}]{\com}\eta\) but also show
that \[\semi[\binte{\low}{\upp}]{\com}\eta = \semi[\inte]{\com}\eta\]
This means that we were able to compute exact interval analysis in
finite time. Then we tries to adopt the same technique to the
non-relational collecting semantics. Indeed one can introduce a
bounded version \(\bbCnr{\low}{\upp}\) of \(\bCnr\) where a \(\top\)
element is used for variables that exceed the bounds. However in this
way we loose too much information to infer equality with the
non-relational collecting semantics.  Therefore we introduced
\(\btbCnr{\low}{\upp}\), bounded non-relational collecting with a
smashed top element. With this analysis we could infer equality up to
the top element, meaning that if the bounded analysis with smashed top
resulted in something different than \(\top\) then its result was the
same as the non-relational collecting, i.e.,
\[\btsemnr{\low}{\upp}{\com}\eta \neq \top \implies \semnr{\com}\eta =
  \btsemnr{\low}{\upp}{\com}\eta\] Intuitively, this implies that we
can decide the termination of non-relational collecting analysis. The
bounded version with the smashed \(\top\) element, only infers this
\(\top\) element if, during the analysis, a variable exceeds the
predefined bounds and follows an infinite sequence. Consequently, the
original non-relational collecting analysis does not terminate under
these conditions using the Kleene iteration. While if something other
than \(\top\) is inferred, it means that it is the most precise
invariant.

\paragraph*{Future work.}
Throughout the development of this thesis, we encountered various
obstacles, which we present here as potential extensions and
directions for future research. Addressing all of these issues was
beyond the scope of this thesis due to time constraints.

The first and most prominent question is whether it is possible to
precisely compute the non-relational collecting semantics. Our
hypothesis is that it is indeed feasible, as suggested by the work
of~\cite{Lefaucheux2024}. Their research builds on Presburger
arithmetic from~\cite{1571698599431503232}, which, although beyond the
scope of this thesis, presents interesting results, particularly its
decidability.

A second follow up question might involve the extension of the
language to support non-linear expressions. Our hypothesis in this
case is that invariants might become undecidable, as we are able to
write diophantine equations. Take for example the following code
snippet
\begin{lstlisting}[language=Imp, caption=Program with diofantine equations, label=code5]
  (*\(\vec{x}\)*) := null_vec_size(n); /* where n is the number of
                            variables in the diofrantine
                            equation */
  while (true) do
    ((*\(x_i\)*)++) + ((*\(x_i\)*)--)
    z = P((*\(\vec{x}\)*)) /* Where P((*{\color{codegreen}\(\vec{x}\)}*)) is a diofantine equation */
\end{lstlisting}
If the non-relational collecting semantics were decidable in a
language that permitted non-linear equations, we could determine the
solution of any Diophantine equation by deciding whether
\(0 \in \semnr{{\com[P]}_d}\). However, this is known to be an
undecidable problem, as demonstrated in~\cite{zbMATH03336816}.

Finally, future work can be focused on the application to real-world
scenarios. Aforementioned tools such as
Astreé~\cite{10.1007/978-3-540-31987-0_3} and
MOPSA~\cite{10.1007/978-3-031-30820-8_37} use the traditional widening
technique, which, while sound, does not infer the most precise
invariant in the domain of intervals. In contrast, we have
demonstrated that it is possible to achieve more precise invariants
while ensuring the termination of the analyzer using the domain of
intervals.
